---
title: "Group_Project"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Set Up

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(corrr)
library(readr)
library(readxl)
```

# Asthma Data Exploration

```{r}
# Load the Asthma dataset
asthma_ca <- read_csv("data/lifetime-asthma-prevalence-by-county-2015_2022.csv")
# Replace the incorrect character
asthma_ca$YEARS <- iconv(asthma_ca$YEARS, from = "latin1", to = "UTF-8")
asthma_ca$YEARS <- gsub("[^0-9-]", "-", asthma_ca$YEARS)

# filter for all ages in 2021-2022
asthma_2021_2022 <- asthma_ca %>% 
  filter(`AGE GROUP` == "All ages", YEARS == "2021-2022") %>%
  rename(County = COUNTY)

#getting rid of other columns
asthma_2021_2022 <- asthma_2021_2022 %>% 
  select(County, `LIFETIME PREVALENCE`)
# Delete row "California"
asthma_2021_2022 <- asthma_2021_2022[-1, ]
```

# Data Additions and Exploration

## Air quality data

```{r}
pm25_ca<-read_csv("data/pm25_CA_22.csv")
ozone_ca<-read_csv("data/ozone_ca.csv")
pm10_ca<-read_csv("data/pm10_ca.csv")

#View(pm25_ca)
#view(no2_ca)
#view(ozone_ca)
#view(asthma_2021_2022)
```

### Temp and precipitation data pull and exploration

```{r}
temp <- read_excel("data/temp.xlsx")
precip <- read_excel("data/precip.xlsx")
#view(temp)

# Filter for California
temp_ca<- temp %>%
  filter(State == "California")

precip_ca <- precip %>%
  filter(State == "California")
  
```

```{r}
# Remove " County" from the County column in temp and precip
temp_ca$Name <- sub(" County$", "", temp_ca$Name)
precip_ca$Name <- sub(" County$", "", precip_ca$Name)
```

### Adding Vehicle and Population data:

```{r}
# Reading population and vehicle data in 
pop_ca <- read_excel("data/co-est2023-pop-06.xlsx")
vehicles_ca <- read.csv("data/dmv.CA.csv")
```

```{r}
#cleaning population
pop_ca <- pop_ca %>%
  mutate(County = sub("^\\.", "", County),                # Remove leading dot
         County = sub(" County, California$", "", County))  # Remove trailing text

# Cleaning Vehicles:
vehicles_ca <- vehicles_ca %>%
  slice(-1:-5) %>%
  filter(Measure.Names == "Total Vehicles")

vehicles_ca <- vehicles_ca %>%
  select(-Measure.Names) %>%
  rename(`Total Vehicles` = Measure.Values)

#standardizing County Name
vehicles_ca <- vehicles_ca %>%
  rename(`County` = Counties)
```

### Assessing county uniformity and reducing county data to only counties for which we have complete data

```{r}
# view(asthma_2021_2022)
```

```{r}
# Get unique counties for temp and precip
unique_temp_counties <- unique(temp_ca$Name)
unique_precip_counties <- unique(precip_ca$Name)

# Get common counties across all datasets, including pop_ca and vehicles_ca
common_counties <- Reduce(intersect, list(unique(pm25_ca$County),
                                            unique(ozone_ca$County),
                                            unique(pm10_ca$County),
                                            unique(asthma_2021_2022$County),
                                            unique_temp_counties,
                                            unique_precip_counties))


# Filter each dataset to include only rows with common counties
pm25_ca <- pm25_ca[pm25_ca$County %in% common_counties, ]
ozone_ca <- ozone_ca[ozone_ca$County %in% common_counties, ]
pm10_ca <- pm10_ca[pm10_ca$County %in% common_counties, ]
asthma_2021_2022 <- asthma_2021_2022[asthma_2021_2022$County %in% common_counties, ]
temp_ca <- temp_ca[temp_ca$Name %in% common_counties, ]
precip_ca <- precip_ca[precip_ca$Name %in% common_counties, ]
```

## Plot by Annual Observations

Looking at observations per county:

```{r}
# Count the number of observations per county for each dataset
pm25_county_obs <- table(pm25_ca$County)
ozone_county_obs <- table(ozone_ca$County)
pm10_county_obs <- table(pm10_ca$County)

# Convert the counts to data frames for easier plotting
pm25_df <- data.frame(County = names(pm25_county_obs), 
                      Observations = as.integer(pm25_county_obs))
ozone_df <- data.frame(County = names(ozone_county_obs), 
                       Observations = as.integer(ozone_county_obs))
pm10_df <- data.frame(County = names(pm10_county_obs), 
                      Observations = as.integer(pm10_county_obs))

# Plot the distribution of observations for PM2.5
ggplot(pm25_df, aes(x = County, y = Observations)) +
  geom_bar(stat = "identity", fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "PM2.5 Observations per County", 
       y = "Number of Observations",
       x = "County",
       caption = "Figure 1: Number of PM2.5 observations in each California county for 2022.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )

# Plot the distribution of observations for PM10
ggplot(pm10_df, aes(x = County, y = Observations)) +
  geom_bar(stat = "identity", fill = "purple", color = "black", alpha = 0.7) +
  labs(title = "PM10 Observations per County", 
       y = "Number of Observations",
       x = "County",
       caption = "Figure 2: Number of PM10 observations in each California county for 2022.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )

# Plot the distribution of observations for Ozone
ggplot(ozone_df, aes(x = County, y = Observations)) +
  geom_bar(stat = "identity", fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(title = "Ozone Observations per County", 
       y = "Number of Observations",
       x = "County",
       caption = "Figure 3: Number of ozone observations in each California county for 2022.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )


```

## Plot by Seasonal Observations

Filtering by season

```{r}
# Alter date column from 'chr' to 'date' for all pollutant
## PM10
pm10_ca$Date <- as.Date(pm10_ca$Date, format = "%m/%d/%Y")

##PM2.5
pm25_ca$Date <- as.Date(pm25_ca$Date, format = "%m/%d/%Y")

## Ozone
ozone_ca$Date <- as.Date(ozone_ca$Date, format = "%m/%d/%Y")


# Wet season: Oct - April
## Filter by dates
pm10_ca_wet <- pm10_ca %>%
  filter(month(Date) %in% c(1, 2, 3, 4, 10, 11, 12))
pm25_ca_wet <- pm25_ca %>%
  filter(month(Date) %in% c(1, 2, 3, 4, 10, 11, 12))
ozone_ca_wet <- ozone_ca %>%
  filter(month(Date) %in% c(1, 2, 3, 4, 10, 11, 12))


# Dry season: May - September
## Filter by dates
pm10_ca_dry <- pm10_ca %>%
  filter(month(Date) %in% c(5, 6, 7, 8, 9))
pm25_ca_dry <- pm25_ca %>%
  filter(month(Date) %in% c(5, 6, 7, 8, 9))
ozone_ca_dry <- ozone_ca %>%
  filter(month(Date) %in% c(5, 6, 7, 8, 9))
```

Looking at observations per county:

```{r}
# Count the number of observations per county for each dataset
pm10_county_obs_wet <- table(pm10_ca_wet$County)
pm10_county_obs_dry <- table(pm10_ca_dry$County)

pm25_county_obs_wet <- table(pm25_ca_wet$County)
pm25_county_obs_dry <- table(pm25_ca_dry$County)

ozone_county_obs_wet <- table(ozone_ca_wet$County)
ozone_county_obs_dry <- table(ozone_ca_dry$County)

# Convert the counts to data frames for easier plotting
pm10_df_wet <- data.frame(County = names(pm10_county_obs_wet), Observations = as.integer(pm10_county_obs_wet))
pm10_df_dry <- data.frame(County = names(pm10_county_obs_dry), Observations = as.integer(pm10_county_obs_dry))

pm25_df_wet <- data.frame(County = names(pm25_county_obs_wet), Observations = as.integer(pm25_county_obs_wet))
pm25_df_dry <- data.frame(County = names(pm25_county_obs_dry), Observations = as.integer(pm25_county_obs_dry))

ozone_df_wet <- data.frame(County = names(ozone_county_obs_wet), Observations = as.integer(ozone_county_obs_wet))
ozone_df_dry <- data.frame(County = names(ozone_county_obs_dry), Observations = as.integer(ozone_county_obs_dry))
```

Combine wet season pollutants into one dataframe Combine dry seaosn pollutants into one dataframe

```{r}
# Add a column to each wet season dataframe to specify the pollutant type
pm10_df_wet <- pm10_df_wet %>% 
  mutate(Pollutant = "PM10")
pm25_df_wet <- pm25_df_wet %>%
  mutate(Pollutant = "PM2.5")
ozone_df_wet <- ozone_df_wet %>% 
  mutate(Pollutant = "Ozone")

# Add a column to each dry season dataframe to specify the pollutant type
pm10_df_dry <- pm10_df_dry %>%
  mutate(Pollutant = "PM10")
pm25_df_dry <- pm25_df_dry %>%
  mutate(Pollutant = "PM2.5")
ozone_df_dry <- ozone_df_dry %>% 
  mutate(Pollutant = "Ozone")

# Combine wet season dataframes into one
wet_season_data <- bind_rows(pm10_df_wet, pm25_df_wet, ozone_df_wet)

# Remove counties that do not have observations for all 3 pollutants
complete_counties <- wet_season_data %>%
  group_by(County) %>%
  summarise(Unique_Pollutants = n_distinct(Pollutant)) %>%
  filter(Unique_Pollutants == 3)

# Check valid counties
valid_counties <- complete_counties %>% 
  pull(County)

# Filter for only counties with all 3 pollutants
wet_season_data <- wet_season_data %>%
  filter(County %in% valid_counties)

# Combine dry season dataframes into one
dry_season_data <- bind_rows(pm10_df_dry,pm25_df_dry,ozone_df_dry)

# Remove counties that do not have observations for all 4 pollutants
complete_counties2 <- dry_season_data %>%
  group_by(County) %>%
  summarise(Unique_Pollutants = n_distinct(Pollutant)) %>%
  filter(Unique_Pollutants == 3)

# Check valid counties
valid_counties2 <- complete_counties2 %>% 
  pull(County)

# Filter for only counties with all 4 pollutants
dry_season_data <- dry_season_data %>%
  filter(County %in% valid_counties2)
```

Figure 4

```{r}
# Distribution on PM10 across counties in the wet season (Jan-April, Oct-Dec)
#ggplot(pm10_df_wet, aes(x=County, y=Observations)) + geom_bar(stat = "identity", fill="lightblue", color = "black", alpha = 0.7) + labs(title = "PM10 Observations by County during the Wet Season", x="Counties", y = "Number of Observations") + theme_minimal()

# Create a combined bar chart
ggplot(wet_season_data, aes(x=County, y=Observations, fill=Pollutant)) +
  geom_bar(stat="identity", position="dodge", color="black") +
  labs(title=" 2022 Pollutant Observations by County during the Wet Season (Jan-April, Oct-Dec)",
       x="Counties",
       y="Number of Observations") +
  theme_minimal() +
  scale_fill_manual(values=c("PM10"="lightblue", "Ozone"="lightgreen", "PM2.5" = "pink")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 5

```{r}
ggplot(dry_season_data, aes(x=County, y=Observations, fill=Pollutant)) +
  geom_bar(stat="identity", position = "dodge", color="black") +
  labs(title="2022 Pollutant Observations by County during the Dry Season (May-Sept)",
       x="Counties", y="Number of Observations") +
  theme_minimal() +
  scale_fill_manual(values = c("PM10"="lightblue", "Ozone" = "lightgreen", "PM2.5" = "pink")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Set a Threshold

```{r}
# Apply a 10% threshold for each pollutant to 
# remove counties with too few observations
ozone_threshold <- quantile(ozone_df$Observations, 0.10)
pm25_threshold <- quantile(pm25_df$Observations, 0.10)
pm10_threshold <- quantile(pm10_df$Observations, 0.10)

# filter datasets
ozone_10percent <- ozone_df %>%
  filter(Observations >= ozone_threshold)

pm25_10percent <- pm25_df %>%
  filter(Observations >= pm25_threshold)

pm10_10percent <- pm10_df %>%
  filter(Observations >= pm10_threshold)

# Find common counties for each pollutants
common_counties_2 <- Reduce(intersect,
                            list(ozone_10percent$County,
                                 pm25_10percent$County,
                                 pm10_10percent$County))

# Filter the full datasets
ozone_ca <- ozone_ca %>% filter(County %in% common_counties_2)
pm25_ca <- pm25_ca %>% filter(County %in% common_counties_2)
pm10_ca <- pm10_ca %>% filter(County %in% common_counties_2)
```

## Merging all 8 datasets into one final dataframe

```{r}
# Calculate mean pollutant levels per county
ozone_summary <- ozone_ca %>% 
  group_by(County) %>% 
  summarise(Ozone = mean(`Daily Max 8-hour Ozone Concentration`, na.rm = TRUE))
pm25_summary <- pm25_ca %>% 
  group_by(County) %>% 
  summarise(PM2.5 = mean(`Daily Mean PM2.5 Concentration`, na.rm = TRUE))
pm10_summary <- pm10_ca %>% 
  group_by(County) %>% 
  summarise(PM10 = mean(`Daily Mean PM10 Concentration`, na.rm = TRUE))

# Merge all 3 pollutant into one dataframe
all_pollutant_summary <- ozone_summary %>%
  inner_join(pm25_summary, by = "County") %>%
  inner_join(pm10_summary, by = "County")

# Rename and edit precip and temp data
temp_ca <- temp_ca %>% 
  rename(County = Name) %>% 
  select(County, `1901-2000 Mean`) %>% 
  rename(Century_Temp = `1901-2000 Mean`)

precip_ca <- precip_ca %>% 
  rename(County = Name) %>% 
  select(County, `1901-2000 Mean`) %>% 
  rename(Century_Precip = `1901-2000 Mean`)

# Merge temperature and precipitation data on the County column
weather_summary <- inner_join(temp_ca, precip_ca, by = "County")

# Now, merge the pollutant summary (all_pollutant_summary) with the weather summary
pollutant_weather_df <- all_pollutant_summary %>% 
  inner_join(weather_summary, by = "County")
```

```{r}
#need to shorten new DFs to just the common counties
pop_ca <- pop_ca[pop_ca$County %in% common_counties_2, ]

common_counties_2 <- toupper(common_counties_2)
vehicles_ca <- vehicles_ca[vehicles_ca$County %in% common_counties_2, ]

#view(pollutant_weather_df)
#view(pop_ca)
#view(vehicles_ca)
```

```{r}
# Ensure County names are standardized in all data frames
pollutant_weather_df <- pollutant_weather_df %>% mutate(County = toupper(County))
pop_ca <- pop_ca %>% mutate(County = toupper(County))
vehicles_ca <- vehicles_ca %>% mutate(County = toupper(County))
asthma_ca_final <- asthma_2021_2022 %>% mutate(County = toupper(County))
asthma_ca_final_1 <- asthma_ca_final[asthma_ca_final$County %in% common_counties_2, ]
```

```{r}
# Merge pop_ca with pollutant_weather_df first, then merge vehicles_ca
final_df <- pollutant_weather_df %>% 
  inner_join(pop_ca, by = "County") %>% 
  inner_join(asthma_ca_final_1, by = "County") %>%
  inner_join(vehicles_ca, by = "County")

# view(final_df)
```

## Pearson's R test of all variables

```{r}
# loop function for each variable:
# Extract only numeric columns (County is excluded)
numeric_vars <- final_df %>% select(where(is.numeric))

# Initialize an empty data frame to store results
corr_results <- data.frame(Var1 = character(),
                           Var2 = character(),
                           Pearson_R = numeric(),
                           P_Value = numeric(),
                           stringsAsFactors = FALSE)

# Loop over each pair of numeric variables
for (i in 1:(ncol(numeric_vars)-1)) {
  for (j in (i+1):ncol(numeric_vars)) {
    var1 <- names(numeric_vars)[i]
    var2 <- names(numeric_vars)[j]
    
    # Run Pearson's correlation test
    test <- cor.test(numeric_vars[[i]], numeric_vars[[j]], method = "pearson")
    corr_results <- rbind(corr_results, data.frame(Var1 = var1,
                                                   Var2 = var2,
                                                   Pearson_R = test$estimate,
                                                   P_Value = test$p.value,
                                                   stringsAsFactors = FALSE))
  }
}

print(corr_results)
```

```{r}
# Find correlation matrix
cor_matrix <- final_df %>%
  select(-County) %>%
  correlate()
print(cor_matrix)
```

The results indicate that all Pearson correlation coefficients (r-values) are below 0.8 EXCEPT population and vehicles. Only population was used in the following regression model.

## Visualizing Pairwise Correlations

```{r}
# Ozone vs. PM2.5
ggplot(all_pollutant_summary, aes(x = PM2.5, y = Ozone)) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  labs(title = "Concentrations of Ozone vs. PM2.5", 
       x = "PM2.5 (µg/m³)",
       y = "Ozone (ppm)",
       caption = "Figure 6. Correlations of ozone and PM2.5 concentrations.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )

# Ozone vs. PM10
ggplot(all_pollutant_summary, aes(x = PM10, y = Ozone)) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  labs(title = "Concentrations of Ozone vs. PM10", 
       x = "PM10 (µg/m³)", 
       y = "Ozone (ppm)",
       caption = "Figure 7. Correlations of ozone and PM10 concentrations.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )

# PM2.5 vs. PM10
ggplot(all_pollutant_summary, aes(x = PM10, y = PM2.5)) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  labs(title = "Concentrations of PM2.5 vs. PM10", 
       x = "PM10 (µg/m³)", 
       y = "PM2.5 (µg/m³)",
       caption = "Figure 8. Correlations of PM2.5 and PM10 concentrations.") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )
```

## For loop to plot variables against Asthma

```{r}
#Variable list
numeric_vars <- c("Ozone", "PM2.5", "PM10", "Century_Temp", 
                  "Century_Precip", "Population")

# For loop for plotting
for (var in numeric_vars) {
  plot(final_df$`LIFETIME PREVALENCE`, final_df[[var]],
       xlab = "LIFETIME PREVALENCE",
       ylab = var,
       main = paste(var, "vs. LIFETIME PREVALENCE"))
}


```

# Regression Models

```{r}
# 3 separate regression models 
lm_PM2.5 <- lm(`LIFETIME PREVALENCE` ~ `PM2.5` + Century_Temp +
                   Century_Precip + Population, data = final_df)
summary(lm_PM2.5)

lm_PM10 <- lm(`LIFETIME PREVALENCE` ~ `PM10` + Century_Temp +
                   Century_Precip + Population, data = final_df)
summary(lm_PM10)

lm_ozone <- lm(`LIFETIME PREVALENCE` ~ `Ozone` + Century_Temp +
                   Century_Precip + Population, data = final_df)
summary(lm_ozone)

# Full model using all predictors
lm_all <- lm(`LIFETIME PREVALENCE` ~ Ozone + `PM2.5` + PM10 + Century_Temp +
                   Century_Precip + Population, data = final_df)
summary(lm_all)
```

# Log transformation 

```{r}
# Copy the original data
final_df_log <- final_df

# Apply log transformation to all independent variables
final_df_log$log_Ozone <- log(final_df$Ozone)
final_df_log$log_PM2.5 <- log(final_df$`PM2.5`)
final_df_log$log_PM10 <- log(final_df$PM10)
final_df_log$log_Temp <- log(final_df$Century_Temp)
final_df_log$log_Precip <- log(final_df$Century_Precip)
final_df_log$log_Population <- log(final_df$Population)

# Run the model
lm_all_log <- lm(`LIFETIME PREVALENCE` ~ log_Ozone + log_PM2.5 + log_PM10 + 
                   log_Temp + log_Precip + log_Population, data = final_df_log)
summary(lm_all_log)
```

# Evaluate Model Diagnostic

```{r}
AIC(lm_all)
AIC(lm_all_log)
plot(lm_all)
plot(lm_all_log)
```


